---
title: "Mixed Effects Modeling Protocol"
output:
  html_document:
    df_print: paged
data: "2026-04-02"
---

```{r}
library(lme4) # standard for mixed effect models 
library(nlme) # alternative 
library(lmertest) # for more functions 
library(ggplot2)

```
# Example One: OWLS
```{r}
 data = read.csv("Owls.csv")
```

<br> 
## First 
We start with a linear model. <br> 
We are interested in Nestling Negotiation as a function of parental sex, arrival time, and food treatment. <br> 

```{r}
mod.lm <- lm(NegPerChick ~ SexParent * FoodTreatment + 
               SexParent * ArrivalTime, data = data)
plot(mod.lm)
```
<br> 
<br> 
__ALWAYS check assumptions__ <br> 
Our residuals are not homescedastic. The shape of the plot implies a log-transformation may be useful. <br> 
<br> 
<br>
Add the new column and run the new model on the log-transformed data. 

 
```{r}
data$logNeg = log10(data$NegPerChick + 1) 
mod2.lm <- lm(logNeg ~ SexParent * FoodTreatment + 
               SexParent * ArrivalTime, data = data)
plot(mod2.lm)
```
<br> 
The log transformation evened the spread of our residuals, so we can now proceed with the model with assumptions non-violated. <br> 


## We are interested in variation across nest, so
We examine the residuals across nest to assess whether nests account for any variation. <br> 
```{r}
E = resid(mod2.lm)
boxplot(E ~ Nest, data = data, axes = T)
abline(0,0)
```
<br> 
By viewing the boxplot, we can see that variation (*error in the model*) is not evenly spread accross all _Nests_: <br> 
- Boxes are not centered on the 0 line <br> 
- Variation in box height <br> 
- Variation in box width <br> 
<br> 
**Thus**: Nests accounts for variation in the model that is not encompassed by our current predictors. <br> 



### Does adding a random effect *improve* the model? 
We can test whether including a random effect improves the model fit. <br> 
<br>
`gls()` is in the `nlme` package <br> 
- It fits a model using Generalized Least Squares <br> 
- lm assumes constant variance and independent errors, gls relaxes these assumptions <br> 
- gls allows for more complex error structures <br> 
<br> 

```{r}
mod.gls <- gls(logNeg ~ SexParent * FoodTreatment + 
               SexParent * ArrivalTime, data = data)

# lme is in the nlme package 
mod1.lme <- lme(logNeg ~ SexParent * FoodTreatment + 
               SexParent * ArrivalTime,
               # separate (1) intercept for each 'Nest'
               random = ~1 | Nest, 
               data = data)

```
<br> 

__Now we COMPARE our two models__<br> 
We compare the two models using a Likelihood Ratio Test <br> 
```{r}
# LRT to test whether we need the random intercept 
anova(mod.gls, mod1.lme)
```
<br> 
We can see that the AIC of the model *including* the random effect is much smaller, meaning the random effect **improves** the model. <br> 
<br> 


## Now we know we *need* to include the Random Effect 
First we examine the significance of our regression parameters: 
```{r}
summary(mod1.lme)
```
With the p-values, we can see that Sex and ArrivalTime have significant effects, but not the interactions <br> 
<br> 
<br> 
Now we can refit the model using Maximum Likelihood instead of REML for LRT <br>
```{r}
mod1.FULL <- lme(logNeg ~ SexParent * FoodTreatment + 
               SexParent * ArrivalTime,
               # separate (1) intercept for each 'Nest'
               random = ~1 | Nest, 
               data = data, method = "ML") # change to ML
```
<br> 
This is our full model, but we want to build *reduced* models that are different by only **one** treatment, to test for the effect of that treatment by LRT. <br> 
<br> 
```{r}
# remove the sex and food treatment interaction
mod1.A <- update(mod1.FULL, .~. -SexParent:FoodTreatment) 

# remove the sex and arrival time interaction 
mod1.B <- update(mod1.FULL, .~. -SexParent:ArrivalTime)
```

<br> 
Now to test the interactions we can run a test comparing the reduced and full models. <br> 
```{r}
anova(mod1.FULL, mod1.A)
anova(mod1.FULL, mod1.B)
```
<br> 
We can see that our interactions were __NOT__ significant, so we want to remove them from our model. 


## Building the Reduced Updated Model 
```{r}
mod2.FULL = lme(logNeg ~ SexParent + FoodTreatment + ArrivalTime, 
                random = ~1 |Nest, method = "ML", data = data) 
mod2.A = update(mod2.FULL, .~. -FoodTreatment)
mod2.B = update(mod2.FULL, .~. -SexParent) 
mod2.C = update(mod2.FULL, .~. -ArrivalTime)
```
<br>
This is our updated model without those *useless* interaction terms. <br>
<br> 
But we want to see if each factor has a significant effect on the model fit and outcome, so we use the LRT to compare the full model to each reduced model (one treatment removed) <br> 
```{r}
anova(mod2.FULL, mod2.A) # test effect of FoodTreatment
anova(mod2.FULL, mod2.B) # test effect of SexParent
anova(mod2.FULL, mod2.C) # test effect of ArrivalTime
```
<br> 







# Example 2: Marine Benthic Data and Plotting Results
Benthic data from 9 intertidal zones along the east coast. 5 samples were taken per beach and assessed for species richness. <br> 
```{r}
data = read.csv("RIKZ.csv")
data$fBeach = factor(data$Beach) # to ensure Beach is a factor
```

## First: Define Primary Model 
```{r}
M1me1 = lme(Richness ~ NAP, random = ~1 | fBeach, data = data)
summary(M1me1)
```

## Plot 
```{r}
F0 = fitted(M1me1, level = 0) # average relationship over ALL data
F1 = fitted(M1me1, level = 1)
I = order(data$NAP) # order the data
NAPs = sort(data$NAP) # sort the data 
```

<br> 
We need to write a `for` loop to summarize all of the relationship per Beach (one different line for each beach) <br> 

```{r}
plot(NAPs, F0[I], lwd = 4, type = "l", ylim = c(0,22), 
     ylab = "Richness", xlab = "NAP")
for(i in 1:9){
  x1 = data$NAP[data$NAP ==i]
  y1 = F1[data$Beach == i]
  K = order(x1) 
  lines(sort(x1), y1[K])
}

```


We need to write a `for` loop to produce line-summaries for each individual beach. <br>

```{r}
for(i in 1:9){
  x1 = data$NAP[data$NAP ==i]
  y1 = F1[data$Beach == i]
  K = order(x1) 
  plot(lines(sort(x1), y1[K]))
}
```
<br> 

When we plot the points with different colors per Beach, we see that the slopes **differ** between beaches, thus we need a more complex model that includes **random slopes** <br> 


### Plot in ggplot
```{r}
M1me2 = lmer(Richness ~ NAP + (1 | fBeach), data = data)
```

__First__ we are going to plot in `ggplot2` so we need to save coefficients <br> 
```{r}
intercepts = coef(M1me2)$fBeach[,1] # intercept per beach
slopes = coef(M1me2)$fBeach[,2] # slope per beach
pop_int = fixef(M1me2)[1] # population intercept (fixed effects)
pop_slope = fixef(M1me2)[2] # population slope (fixed effects )
```


Now with our variables we can plot in `ggplot2` <br> 
```{r}
ggplot(data, aes(NAP, Richness)) + 
  geom_point(data = data, mapping = aes(color = fBeach)) + 
  geom_abline(intercept = intercepts, slope = slopes, size = 0.25) + 
  geom_abline(intercept = pop_int, slope = pop_slope, size = 1, color = "steelblue") + theme_classic()
```

`ggplot2` produces roughly the same plot (just slightly easier) <br> 




##  More appropriate (Random Slope) mixed effects model 
We saw that the slope of the points differed depending on what Beach they were sampled from, so we should incorporate __Random Slopes__ in our model as well <br> 
```{r}
M1me3 = lmer(Richness ~ NAP + (1 + NAP | fBeach), data = data)
```

We incorporated random *slopes* for by adding `NAP` in our random effect: `(1 + NAP | fBeach)` <br> 
<br> 

Now we can redo our plot to assess the model fit. <br> 
```{r}
intercepts = coef(M1me3)$fBeach[,1] # intercept per beach
slopes = coef(M1me3)$fBeach[,2] # slope per beach
pop_int = fixef(M1me3)[1] # population intercept (fixed effects)
pop_slope = fixef(M1me3)[2] # population slope (fixed effects )

ggplot(data, aes(NAP, Richness)) + 
  geom_point(data = data, mapping = aes(color = fBeach)) + 
  geom_abline(intercept = intercepts, slope = slopes, size = 0.25) + 
  geom_abline(intercept = pop_int, slope = pop_slope, size = 1, color = "steelblue") + theme_classic()

```

__Now__ we see that each Beach's richness changes at different rates. 





# Example 3: kluane data 
```{r}
x = read.csv("kluane.csv")
x
```

Data has a variety of different plots, and different treattments across plots. Thus, we have non-independence between sample units, and need to incorporate a __block design__. <br> 
<br> 

## First we explore the dataset in general 
We want to explore the pattern of the residuals in our model to assess for homoscedasticity (an equal amount of spread per group across the x axis) <br> 
<br> 
Unequal variances are followed up with a log-transformed model.
```{r}
z.int1 = lmer(phen.ach ~ treatment + duration + treatment:duration + 
                (1 | plot), data =x)
plot(z.int1) 

z.int2 = lmer(log(phen.ach) ~ treatment + duration + treatment:duration + 
                (1 | plot), data = x)
plot(z.int2)
```

<br> 
The log-transformed data satisfy our assumptions, so we can proceed with it's interpretation: 
```{r}
summary(z.int2) 
anova(z.int2)
```
<br> 
<br> 
## Plotting the data 
Now with our anova (type 3) table, we have our effects and we can make a plot. <br>
```{r}
ggplot(data = x, aes(x = duration, y = log(phen.ach), 
                      fill = duration, colour = duration)) + 
  geom_line(aes(group = plot), col = "black") + 
  geom_point(size = 3, show.legend = F) + 
  facet_wrap(~treatment, nrow = 1) + 
  labs(x = "Duration", y = "log phenolics concentration") + 
  theme_classic()

```





# Example 4: Starling data 
```{r}
dataf # object pulled from the 'starling.Rda'
```

We have masses of individuals across different months, and whether they were in a tree. <br> 
<br> 
Here we are interested in mass changes of individuals over different months. <br> 

## Preliminary plot (with ggplot2) 
```{r}
## group roost sites in facets 
ggplot(dataf, aes(x = mnth, y = stmass)) + 
  geom_point() + 
  geom_line(aes(group = subject)) + 
  facet_grid(~roostsitu) + 
  theme_classic()


## OR color points and lines by roost-site 
ggplot(dataf, aes(mnth, stmass, color = roostsitu)) + 
  geom_point() + 
  geom_line(aes(group = subject)) + 
  theme_classic()
```


## RUN the mixed model 
Since we are running a __Repeated Measures__, we hold `subject` as a random factor 

```{r}
lmer2 <- lmer(stmass~ mnth*roostsitu+(1|subject), data = dataf)
```


### Overlay results from our model on the plot of the raw data
First we need to extract our predicted values. First we will calculate the population level predictions, then we calculate the predictions for all random effects. 
```{r}
dataf$pred1 <- predict(lmer2, re.form = NA) # this ignores random effects 
dataf$pred2 <- predict(lmer2)
```


Now that we have our *predicted* results, we can plot them over our actual results. 
```{r}
# save the faceted observed values as a base plot
g0 = ggplot(dataf, aes(x = mnth, y = stmass)) + 
  geom_point() + 
  geom_line(aes(group = subject)) + 
  facet_grid(~roostsitu) + 
  theme_classic()
print(g0)

# add in the predicted values here 
g0 + 
  # overlay THIN lines for individual predicted responses
  geom_line(color = "gray55", aes(y = pred2, grouop = subject), size = 0.15) + 
  # thicker blue line for population-level predictions
  geom_line(color = "red", aes(y = pred1, group = subject))

```
<br> <br>

If we are interested in the individual predicted responses, we can investigate those by removing the raw data and population-level predictions from the plot. 
```{r}
g0 = ggplot(dataf, aes(mnth, stmass)) +
  theme_classic() 
  
g0 + geom_line(color = "black", aes(y = pred2, group = subject), size = 0.15)
```




# Example 5: Linear mixed effect models for calculating repeatability 
Here, this is essentially a test of *consistency*. Often valuable for assessing accuracy of instruments and/or differences in procedure (i.e. individuals). <br><br> 
Calculates an **interclass correlation coefficient** to be able to assess for differences in precision across measurements. 
<br> <br> 

## Quick plots for visual assessment
```{r}
flycat = read.csv("flycatcher.csv") 

stripchart(patch ~ bird, data = flycat, vertical = T, pch = 16, cex = 1.5, 
           las = 2, xlab = "Individual Bird", ylab = "Patch length (mm)")
```

With our stripchart, we see that the two points per bird generally follow the same pattern across patch length. This indicates repeatability. <br> 


Another way to plot this is with ggplot
```{r}
ggplot(flycat, aes(y = patch, x = factor(year))) + 
  geom_point(size = 3, col = "firebrick", alpha = 0.5) + 
  geom_line(aes(group = bird)) + 
  labs(x = "Year", y = "Patch length (mm)") + 
  theme_classic()
```

Here we want to look at the position of the points **relative** to all other points. <br> 
Essentially looking at the rank order within each factor (`year`). <br> <br> 

## Running the *simple* mixed effects model 
```{r}
z = lmer(patch ~ 1 + (1 | bird), data = flycat)
```

We include the `~1` to only investigate the *universal* intercept, and what variance is attributable to `bird` and pure error. <br> <br> 

```{r}
VarCorr(z)

# calculate the intraclass correlation coefficient 
r = 1.11504^2/(1.11504^2 + 0.59833^2) 
# variation by random factor (`bird`) divided by total variation
r
```

`VarCorr` gives the amount of variance explained by `trait` (in this case, `bird`). <br> 

