---
title: "Randomization, Bootstraping, and Monte Carlo Methods"
output: html_notebook
date: "2026-04-14"
---

Randomization tests are useful because they do not assume **any** underlying distribution. <br> 


# Basic built-in Functionality 
Sampling from distributions can be useful. 
```{r}
# sample normal distribution
rand.norm = rnorm(n = 10, mean = 0, sd = 1) 
```

## Sampling Distributions
To ensure its a normal distribution, we can increase our sample size (Central Limit Theorem)
```{r}
rand.norm.1000 = rnorm(n = 1000, mean = 0, sd = 1) 
hist(rand.norm.1000)
```


We can also sample other distributions. The uniform distribution is helpful when we want to sample values in a range with each value being equally likely. 
```{r}
rand.unif = runif(n = 1000, min = 0, max = 1)
hist(rand.unif)
```

Bars are similar heights due to a similar frequency of each being sampled. Thus, each value in the range has an *equal chance* of being selected. <br> <br> 

A __Poisson Distriubtion__ is very helpful in biology because many processes can be fit with the 'count data' distribution of a Poisson. <br> 
A poisson process describes events happening with some probability over space or time. Lambda is both the mean and variance of the distribution. 
```{r}
rand.pois = rpois(n = 1000, lambda = 1) 
hist(rand.pois)
```

## Sampling from a Vector 
Often, you want to be able to sample from a prescribed set of values (a vector).
```{r}
# sample a vector 
rand.numb.1 = sample(x = 1:10, size = 1) |> print()
```
<br><br> 

We can also define the vector first, and then sample from it. 
```{r}
x = c(1:10)
z = sample(x, size = 10) |> print()
```
<br><br>
We can also define replacement (values reinputted to vector if previously sampled) or not. The default in sample is `replace = F`. 
```{r}
# sample vector with replacement 
rand.numb.r = sample(x, 10, replace = T) |> print()
```
<br> 


## Weighted Sampling 
If certain values are more probable than others within the vector, we can specify the exact probability of each value in the vector. This requires a _probability vector_ <br> 
```{r}
# establish a probability vector 
prob.vec = c(rep(x = 0.05, times = 5), rep(x = 0.15, times = 5)) |> print()
sum(prob.vec) |> print()
```
<br><br> 

Now we bias our sampling based off of this probability vector. Each value in the probability vector lines up to its coresponding value in the sample vector's position. 
```{r}
# weighted sampling 
rand.numb.bias = sample(x, size = 10, replace = T, prob = prob.vec) |> print()
```
<br> 
Note that the results are dominated by values in the vector >10. This is because our last 5 values in the probability vector ar three times as large as the probabilities for the first 5 values. 




# Randomization for Hypothesis Testing 
Many

```{r}
birds = read.csv("Bumpus_data.csv")
```
<br>
With our data, we want to test the hypothesis that _surviving sparrows_ will have a lower mass than _parishing sparrows_. 
<br> 
This *could* easily be done with a Student's t-test, a simple difference between two independent groups: 
```{r}
bird.t.test = t.test(birds$wgt ~ birds$surv, alternative = "less") |> print()
```
<br>
Our dataset is large enough that we do __not__ have to be concerned with violating the assumptions of a t-test. However, this may not always be the case. <br><br> 

With resampling/randomization test, we are able to relax assumptions of a t-test (particularly useful in small sample sizes). 
<br> 
Randomization allows us to build a probability distrubition by continually resampling __our__ dataset, and finding out how likely our __initial__ finding is given this probability distribution. 
```{r Randomization: for-loop}
iter <- 9999;  # This is the amount of iterations/runs we want 
diff <- NULL;  # A vector to store differences between groups in each iteration
N_birds <- dim(birds)[1] # Total number of birds 

for(i in 1:iter){
  bird_samp <- sample(x = birds$surv, size = N_birds, replace = F) 
  samp_alive <- which(bird_samp == "alive") 
  samp_dead <- which(bird_samp == "dead") 
  mu_samp_A <- mean(birds[samp_alive, 5])
  mu_samp_D <- mean(birds[samp_dead, 5])
  diff[i] <- mu_samp_A - mu_samp_D
}

tru_w_A <- mean(birds[birds$surv == "alive", 5]) 
tru_w_D <- mean(birds[birds$surv == "dead", 5])
td <- tru_w_A - tru_w_D
hist(diff)
abline(v = td, col = "red")
```
<br> 
The randomization generates a **Null** distribution of differences in body weights. Our distribution assumes that there is __no association__ of body weight and survival status. <br><br> 

Now we can find the likelihood of observing our true value given this null distribution.  
```{r}
less_equal_to <- diff[diff <= td] # values less than or equal to observed 
(length(less_equal_to)+1)/(length(diff)+1) # proportion of these values to the total 

# we add 1 due to our observed value being included
```
<br> 

With our randomization test, our observed result is only 0.0059% likely given our simulated null distribution. This is our simulated p-value. 






## Spatial Randomization 
```{r}
fpet <- read.csv("FpetTrees.csv")
```
We want to plot these trees as points in geographic space. 

```{r}
lat1 = 28.289
lat2 = 28.2918
lon1 = -113.1145
lon2 = -113.1095

plot(x = fpet[,4], y = fpet[,3], xlab = "Longitude", ylab = "Latitude", 
     pch = 20, cex.lab = 1.25, cex.axis = 1.25, 
     xlim = c(lon1, lon1-0.001, lon2 + 0.001), 
     ylim = c(lat1 - 0.0005, lat2 +0.0005))
polygon(x = c(lon1, lon1, lon2, lon2)), y = c(lat1,lat2,lat2,lat1), col = "tan")
points()
```

```{r get mean nearest neightbour}
mean_nearest_neighbour <- function(trees, lat = 3, long = 4){ 
  N_trees = dim()}
```






## Randomization for Relationshiops
Correlations and regressions assume a normal distribution and a linear relationship. <br><br> 
This may not always be the case for our data. 

```{r Randomization Relationship Data}
nitrogen <- seq(10,50, length = 30) 
biomass <- 5 + 0.8 * nitrogen + rnorm(30, sd = 8)
plot(biomass~nitrogen)
```
<br> 
First let's look at the actual correlation between our variables. <br> <br> 
```{r}
obs_r = cor(nitrogen, biomass) |> print()
```
<br> <br> 
A correlation works well with tight fitted data, but there are often outliers in true observed data, especially in ecology.  
```{r}
biomass[5] <- 80
plot(biomass~nitrogen)
```
<br> 
Removing outliers to deal with their influence can remove true variation in the data that we are interested in. This is where randomization comes in. <br><br> 
Randomization allows us to incorporate all of the data, even egregious outliers. <br> 
```{r Randomization test for Relationship }
reps = 9999
null_rs <- numeric(reps)
N_biomass <- length(biomass) 

for(i in seq_len(reps)){
  # First, we shuffle the variables to break apart any actual relationship
  shuffled_biomass <- sample(biomass, size = N_biomass, replace = F)
  
  # Now we calculate the correlation coefficient for this run 
  null_rs[i] <- cor(nitrogen, shuffled_biomass)
}

hist(null_rs)
abline(v = obs_r, col = "red")
```
<br> 
We can see that our observed correlation coefficient (the red line) is well outside the 95% confidence interval. 
<br>
```{r Caluclate our p-value}
p.value <- (sum(abs(null_rs) >= abs(obs_r)) / (reps + 1)) |> print()
```

__The Outlier Problem:__<br>
Standard correlation tests are very sensitive to outliers. Randomization tests are based on the actual distribution of your specific data, making them less likely to give a "false significance" (type I error) driven by a *single* datapoint. <br><br> 



## Avoidance/Co-occurence Tests 
We have present (1) and absent (0) data. We want to test if two species show up together more than we would expect due to chance. Do they prefer to cohabitate? 

```{r Set up the data}
# 50 plots 
n_plots <- 50

# Generate random occurrence data for both species 
woodpecker <- rbinom(n_plots, 1, 0.4) # assign "1" with "40%" probability
nuthatch <- rbinom(n_plots, 1, 0.4)

# Force a few extra aco-occurrences to simulate a relationshjip 
woodpecker[1:10] <- 1
nuthatch[1:10] <- 1

# Put into a matrix 
site_mat <- data.frame(woodpecker, nuthatch)
```
<br><br> 
Now we set up our test statistic we will use, and we calculate it for our *true* observed value. <br> 
```{r Test statistic}
obs_cooccur <- sum(site_mat$woodpecker == 1 & site_mat$nuthatch == 1) |> print()
```
<br> 
This is simply the number of sites the two species co-occur in. <br><br> 
Now we set up our randomization test/model. 
```{r}
reps <- 9999
null_counts <- numeric(reps)

for(i in seq_len(reps)){
  # shuffle the nuthatch sites
  shuffled_nuthatch <- sample(site_mat$nuthatch) 
  
  #shuffle the woodpecker sites 
  shuffled_woodpecker <- sample(site_mat$woodpecker) 
  
  # calculate frequency of co-occurrence 
  null_counts[i] <- sum(shuffled_woodpecker == 1 & shuffled_nuthatch == 1) 
}
```
<br><br> 

Now we can make a *pretty* plot. 
```{r}
ggplot() + 
  ylab("Count") +
  xlab("Simulated Co-Occurrence") + 
  geom_histogram(aes(x = null_counts), bins = 30, 
                 fill = "steelblue", alpha = 0.4, color = "black") + 
  geom_vline(xintercept = obs_cooccur, size = 1, 
             linetype = "dashed", color = "red") + theme_classic()
```
<br><br> 

Now we calculate our p-value for co-occurrence:
```{r Calculating p-value}
p.value <- (sum(null_counts >= obs_cooccur)/(reps + 1)) |> print() 
```

__Chi-Squared Limitations:__ <br> 
In ecology, some species are naturally rare and others can be very _common_. The standard use of a Chi-squared test can be invalid when species are common in general. By shuffling our locations but keeping the total count fixed, you account for the fact that a common species is likely to overlap with others just by being everywhere. <br><br> 


## ANOVA-design Randomization (>2 groups)
A one-way ANOVA is very useful for comparing groups. However, ecological data--such as frequency of species occurrence--is often *count* data, and thus does not follow a normal distribution, but a *Poisson* distribution. The 0's in a Poisson violate the assumptions of an ANOVA. <br><br> <br>
__The example:__ <br>
We want to test three different methods to restore native prairie: controlled burning, conservation grazing, and mowing (control). We measured the density of native grass stems per square meter in 15 plots for each treatment (45 plots total). <br><br> 
<br> 
a __Randomization ANOVA__ allows us to test for differences accross mulitple groups without worrying about whether your residuals are perfectly bell-shaped. <br><br> 
```{r Generate our example data}
burning <- rpois(15, 25) # mean of 25 stems 
grazing <- rpois(15, 18) # mean of 18
mowing <- rpois(15, 12) # mean of 12

restoration_data <- data.frame(
  stems = c(burning, grazing, mowing), 
  treatment = rep(c("Burn","Graze","Mow"), each = 15)
) |> print()
```
Now we have all of our grass data for the three treatments in one dataframe. <br><br>
Now, we decide on, and calculate our test statistic. Since it is an ANOVA design, we will use the **F statistic**. <br><br> 

```{r Our observed F-statistic}
obs_anova <- summary(aov(stems ~ treatment, data = restoration_data))
obs_F <- obs_anova[[1]]["treatment", "F value"]
```
<br><br>
Now we can randomize our data and simulate a distribution of F statistics. To do this, we can simply shuffle **group membership** instead of the dependent variable data itself. This shuffles the assignment just the same. 
```{r}
reps <- 4999 
null_f_values <- numeric(reps)

for(i in seq_len(reps)){ 
  # Shuffle treatment
  shuffled_labeels <- sample(restoration_data$treatment) # NO replacement
  
  # Run ANOVA on the shuffled data and grab the F values
  s_anova <- summary(aov(restoration_data$stems ~ shuffled_labeels))
  null_f_values[i] <- s_anova[[1]]["shuffled_labeels", "F value"]
}

hist(null_f_values)
abline(v = obs_F, col = "red")
```
Notice that our line for our observed value is not on the histrogram, that is because it is so much larger than what was simulated with our randomization. <br><br>


Now we can calculate the p-value: 
```{r}
p.value = (sum(null_f_values >= obs_F)/ (reps + 1)) |> print() 
```


